---
title: "Getting Started with rmet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rmet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center",
  dpi = 150,
  warning = FALSE,
  message = FALSE
)
```

Brazil's National Institute of Meteorology (INMET) maintains a network of
automatic weather stations that record hourly observations across the country.
The raw data are distributed as annual ZIP archives containing CSV files, one
per station, with formatting inconsistencies that vary across years (e.g., different
date formats, trailing semicolons, phantom columns, and mixed encodings).
`rmet` automates the full pipeline: downloading those archives, resolving
their structural quirks, and returning a clean, analysis-ready `data.frame`.

## Installation

```{r install, eval = FALSE}
# install.packages("devtools")
devtools::install_github("rodrigosqrt3/rmet")
```

```{r library}
library(rmet)
```

## Step 1 — Browse the station catalogue

The package ships with a bundled catalogue of all automatic INMET stations.
`inmet_stations()` lets you filter by state, so you can confirm station codes
before downloading anything.

```{r stations, eval = FALSE}
stations <- inmet_stations(state = "RS")
head(stations[, c("code", "name", "latitude", "longitude", "elevation")])
```

Station **A801** is the Porto Alegre automatic station, located at roughly
−30.05 °S, −51.17 °W, 46.97 m above sea level.

## Step 2 — Download annual archives

`inmet_download()` fetches the annual ZIP files from INMET's servers and saves
them to a persistent local cache. Downloads are resumable: if a connection
drops, re-running the same call picks up where it left off.

```{r download, eval = FALSE}
# Downloads the 2023 archive (~70 MB) to the default cache directory.
# Safe to re-run — skips files that are already complete.
inmet_download(2023)
```

You can inspect what is already cached at any time:

```{r cache, eval = FALSE}
inmet_cache_status()
```

## Step 3 — Read data into R

`inmet_read()` parses the ZIP archives directly — no manual extraction needed.
It normalises column names, converts all measurement columns to `numeric`,
parses timestamps, and shifts UTC to local Brazilian time.

```{r read, eval = FALSE}
df <- inmet_read(
  years      = 2023,
  stations   = "A801",
  start_date = "2023-01-01",
  end_date   = "2023-12-31"
)
```

The examples below use the package's built-in sample dataset, which contains
one year of hourly observations from station A801 (Porto Alegre, 2023).

```{r load-data}
df <- rmet_example
str(df[, c("datetime", "station_code", "state",
           "temp_dry_c", "precip_mm", "humid_rel_pct",
           "wind_speed_ms", "pressure_station_hpa")])
```

## Exploring the data

### Daily aggregation

Hourly data are often most useful after aggregating to daily summaries.

```{r daily-agg}
df$date <- as.Date(df$datetime)

daily <- aggregate(
  cbind(temp_dry_c, precip_mm, humid_rel_pct, wind_speed_ms) ~ date,
  data = df,
  FUN  = function(x) mean(x, na.rm = TRUE)
)

daily$temp_max <- aggregate(temp_dry_c ~ date, df, max, na.rm = TRUE)$temp_dry_c
daily$temp_min <- aggregate(temp_dry_c ~ date, df, min, na.rm = TRUE)$temp_dry_c
daily$precip   <- aggregate(precip_mm  ~ date, df, sum, na.rm = TRUE)$precip_mm
daily$month    <- as.integer(format(daily$date, "%m"))
```

### Temperature series

```{r plot-temp, fig.cap = "Daily temperature range at Porto Alegre (A801) — 2023."}
col_max  <- "#E74C3C"
col_min  <- "#3498DB"
col_band <- "#F1948A"

par(
  mar     = c(4, 4.5, 3, 1),
  family  = "sans",
  cex.lab = 0.95,
  cex.axis = 0.85
)

plot(
  daily$date, daily$temp_max,
  type = "n",
  ylim = range(c(daily$temp_min, daily$temp_max), na.rm = TRUE),
  xlab = "",
  ylab = "Temperature (\u00b0C)",
  main = "Daily Temperature Range \u2014 Porto Alegre (A801), 2023",
  axes = FALSE
)

# shaded band between min and max
polygon(
  c(daily$date, rev(daily$date)),
  c(daily$temp_max, rev(daily$temp_min)),
  col    = adjustcolor(col_band, alpha.f = 0.35),
  border = NA
)

lines(daily$date, daily$temp_max, col = col_max, lwd = 1.2)
lines(daily$date, daily$temp_min, col = col_min, lwd = 1.2)

axis(1,
     at     = seq(min(daily$date), max(daily$date), by = "2 months"),
     labels = format(seq(min(daily$date), max(daily$date), by = "2 months"), "%b"),
     las    = 1)
axis(2, las = 1)
box(col = "grey80")

legend(
  "topright",
  legend = c("Daily max", "Daily min"),
  col    = c(col_max, col_min),
  lwd    = 2,
  bty    = "n",
  cex    = 0.85
)
```

### Precipitation

```{r plot-precip, fig.cap = "Daily accumulated precipitation at Porto Alegre (A801) — 2023."}
col_prec <- "#3498DB"

par(mar = c(4, 4.5, 3, 1), family = "sans", cex.lab = 0.95, cex.axis = 0.85)

barplot(
  daily$precip,
  col    = adjustcolor(col_prec, alpha.f = 0.75),
  border = NA,
  space  = 0,
  xlab   = "",
  ylab   = "Precipitation (mm)",
  main   = "Daily Precipitation \u2014 Porto Alegre (A801), 2023",
  axes   = FALSE
)

axis(2, las = 1)

# Month labels on x axis
month_starts <- which(!duplicated(format(daily$date, "%m")))
axis(1,
     at     = month_starts,
     labels = format(daily$date[month_starts], "%b"),
     tick   = FALSE)

box(col = "grey80")
```

### Monthly temperature boxplots

```{r plot-seasonal, fig.cap = "Seasonal cycle of hourly dry-bulb temperature at Porto Alegre (A801) — 2023."}
col_box <- "#E74C3C"

df$month <- as.integer(format(df$datetime, "%m"))
month_labels <- c("Jan","Feb","Mar","Apr","May","Jun",
                  "Jul","Aug","Sep","Oct","Nov","Dec")

par(mar = c(4, 4.5, 3, 1), family = "sans", cex.lab = 0.95, cex.axis = 0.85)

boxplot(
  temp_dry_c ~ month,
  data     = df,
  col      = adjustcolor(col_box, alpha.f = 0.40),
  border   = col_box,
  names    = month_labels,
  xlab     = "Month",
  ylab     = "Dry-bulb temperature (\u00b0C)",
  main     = "Temperature Seasonality \u2014 Porto Alegre (A801), 2023",
  outline  = FALSE,
  whisklty = 1,
  medlwd   = 2,
  axes     = FALSE
)

axis(1, at = 1:12, labels = month_labels, las = 1)
axis(2, las = 1)
box(col = "grey80")
```

### Wind speed distribution

```{r plot-wind, fig.cap = "Distribution of hourly wind speed at Porto Alegre (A801) — 2023."}
col_wind <- "#2C3E50"

ws <- df$wind_speed_ms[!is.na(df$wind_speed_ms) & df$wind_speed_ms >= 0]

par(mar = c(4, 4.5, 3, 1), family = "sans", cex.lab = 0.95, cex.axis = 0.85)

h <- hist(ws, breaks = 30, plot = FALSE)

plot(h,
     col    = adjustcolor(col_wind, alpha.f = 0.55),
     border = "white",
     xlab   = "Wind speed (m/s)",
     ylab   = "Frequency",
     main   = "Wind Speed Distribution \u2014 Porto Alegre (A801), 2023",
     axes   = FALSE)

abline(v   = mean(ws),
       col = "#E74C3C",
       lwd = 2,
       lty = 2)

axis(1, las = 1)
axis(2, las = 1)
box(col = "grey80")

legend("topright",
       legend = paste0("Mean: ", round(mean(ws), 1), " m/s"),
       col    = "#E74C3C",
       lwd    = 2,
       lty    = 2,
       bty    = "n",
       cex    = 0.85)
```

## Filtering and subsetting

`inmet_read()` supports filtering at read time, which avoids loading
unnecessary data into memory.

```{r filter, eval = FALSE}
# Only summer months (December–February), temperature and humidity
summer <- inmet_read(
  years      = 2023,
  stations   = "A801",
  start_date = "2023-12-01",
  end_date   = "2023-12-31",
  variables  = c("temp_dry_c", "humid_rel_pct")
)

# Multiple stations across Rio Grande do Sul
rs_stations <- inmet_read(
  years    = 2023,
  stations = c("A801", "A802", "A827"),
  variables = c("temp_dry_c", "precip_mm")
)
```

## Extracting CSV files

If you need the raw CSV files for use in other tools, `inmet_extract()` copies
them out of the ZIP archives to a directory of your choice.

```{r extract, eval = FALSE}
inmet_extract(
  years   = 2023,
  out_dir = file.path(tempdir(), "inmet_csv")
)
```

## Cache management

```{r cache-mgmt, eval = FALSE}
# See what is cached and how large each file is
inmet_cache_status()

# Remove all cached archives to free disk space
inmet_cache_clear()
```

## Variable reference

The table below lists all meteorological columns returned by `inmet_read()`.

| Column | Description | Unit |
|---|---|---|
| `temp_dry_c` | Dry-bulb (air) temperature | °C |
| `temp_max_c` | Maximum temperature (previous hour) | °C |
| `temp_min_c` | Minimum temperature (previous hour) | °C |
| `temp_dew_c` | Dew-point temperature | °C |
| `precip_mm` | Total precipitation | mm |
| `pressure_station_hpa` | Atmospheric pressure at station level | hPa |
| `pressure_max_hpa` | Maximum pressure (previous hour) | hPa |
| `pressure_min_hpa` | Minimum pressure (previous hour) | hPa |
| `humid_rel_pct` | Relative humidity | % |
| `humid_rel_max_pct` | Maximum relative humidity (previous hour) | % |
| `humid_rel_min_pct` | Minimum relative humidity (previous hour) | % |
| `wind_speed_ms` | Wind speed | m/s |
| `wind_dir_deg` | Wind direction | degrees |
| `wind_gust_ms` | Wind gust speed | m/s |
| `radiation_kjm2` | Global solar radiation | kJ/m² |

## Data source

All data are provided by INMET and are freely available at
<https://portal.inmet.gov.br/>. The package does not redistribute raw data; it
only automates retrieval and parsing of files hosted by INMET.
